Learning with misspecified models: Overconfidence and Stereotypes

### 2023-10-11 Paper/draft.Rmd
intro start

bibliography: references.bib
fontsize: 12pt 
geometry: margin=2.5cm
---

@@

Overconfidence has been documented in a wide range of settings and has been shown to have important economic consequences:
@Camerer1999 find that overconfidence can explain the excessive entry of entrepreneurs into the market; @Hoffman2020 show that
workers who overestimate their productivity within a firm are less likely to look for a new job; @Oster2013 determine that 
people at high risk of Huntington's disease make saving decisions as if they were not at risk.

Although the bias has been widely studied by economists, the settings in which 
the the belief updating procedure has been studied is narrow in scope. In particular, the literature has focused on
the cases in which the agent updates their beliefs about a single characteristic that they might care about.
For instance, @Eil2011 study the case in which the agent updates their beliefs about their own ability, and their 
own beauty in two separate experiments. Similarly, @Benoit2015 study the case in which the agent updates their 
beliefs about their own IQ.

@@

@@

stargazer(reg, reg_ego, reg_stereotype, reg_bayesian,  
          title='Regression of effort change on news difference with robust standard errors', 
          dep.var.labels='Change in effort', 
          covariate.labels=c('Good news', 'News difference', 'News difference * Good news'),
          column.labels=c('All', 'Ego-relevant', 'Stereotype', 'Bayesian Simulation'), 
          omit.stat=c('f', 'ser'), 
          digits=2, 
          label='tab:regression', 
          header=FALSE, 
          se = list(coefs[, 2], coefs_ego[, 2], coefs_stereotype[, 2], coefs_bayesian[, 2]))

>>>>

bibliography: references.bib
fontsize: 12pt 
geometry: margin=2.5cm

---

@@

A growing body of literature in economics studies the mechanisms through which economic agents come to 
hold incorrect beliefs about a fundamental. Most of this literature focuses on the case in which agents
are passive observers of the world and incorrectly incorporate the information they receive into their 
beliefs\footnote{See \citet{benjamin2019} for a review of error in probabilistic thinking}.
However, in many situations, agents are active participants in the process of generating information.
In such cases, the information they receive is affected by their actions and the way they process it
can affect their future behavior. I refer to this as an \emph{endogenous information} setting.
This sort of setting, has been shown to explain pehnomena that
cannot be rationalized with the classical models of biased belief updating (e.g. @Nyarko1991 and 
@Levy2022). Settings with endogenous information and misspecified learning also offer alternative explanations 
for why certain well known biases might persist. I focus on a group of theories that aim to explain 
the prevalence of overconfidence.

Overconfidence has been shown to be a prevalent bias in a variety of settongs. @Oster2013 shows that 
subjects who are at high risk of having Huntingtons disease underestimate their probability of 
having the disease and make retirement decisions as if they were not sick. Relating to investment 
decisions, @Hoffman2020 show that workers overestimate their productivity in their 
current employment and are unlikely to look for other oportunities. @Camerer1999 find that 
entrepreneurs are overconfident about the quiality of their enterprise, 
which leads to excessive entry and early exit from markets. These are only some examples of how 
holding an incorrect belief about a fundamental can lead to costly decisions \footnote{Other examples 
of overconfidence are }.

I isolate the main forces from 3 different theories of misspecified learning. Each of the theories 
rationalizes the prevalence of overconfidence in frameworks where the agents are learning about 
an ego-relevant parameter and an exogenously determined state. The interaction between these two 
parameters together with an endogenous information process give rise to two explanaitions
for the presistence of overconfidence. On one hand, even with bayesian updating, the agent 
may fall into learning traps. Learning traps are misspecified beliefs and optimal actions 
that induce signals that confirm the misspecified belief. Once an agent finds themselves in 
such a situation, the belief will be stable and even with a correctly speficified model of learning, 
they will not be able to abandon their icorrect beliefs. On the other hand a setting with two variables allows the agent to engage in asymmetric updating 
with respect to good and bad news. When the outcome depends on an ego-relevant parameter as well
as a state, the agent can attribute good news to a high value of the parameter while attributing 
failures to a low value of the exogenous parameter\footnote{This is in line with the well known theory 
in behavioral science of self-attribution bias of which \citet{kelley1980} provides a review.}.

@Heidhues2018 shows 
that under certain assumptions on the output function, learning traps can rationalize the 
existance of misspecified beliefs when agents are dogmatic about the ego-relevant parameter. 
In a more general model @Ba2023 further analyzes this setting 
and characterizes the stuations in which even agents who are not dogmatic might become trapped.
She allows initi dogmatic agents to consider alternative theories and adopt whichever is a 
better explanation of the observed data. They do so by performing Bayesian hypothesis tests
that compare the likelihood of having observed the outcome under alternative paradigms\footnote{A 
similar mechanism is proposed by \citet{Schwarstein2021} in a setting with persuasion}. Although
this learning mechanism allows agents to avoid the learning traps, there are still situations under
which misspecified beliefs may persist.

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
dogmatic <- updates[, .(round_number, good_news_dogmatic, signal_dogmatic_diff, effort_change_dogmatic)]
setnames(dogmatic, c('good_news_dogmatic', 'signal_dogmatic_diff', 'effort_change_dogmatic'), c('good_news', 'news_diff', 'effort_change'))

# run the regression for the dogmatic simulation
reg_dogmatic <- lm(effort_change ~ good_news + news_diff + news_diff:good_news, data=dogmatic[round_number>1])

# Attach the robust standard errors to the coefficient estimates
coefs_dogmatic <- coef(reg_dogmatic)
coefs_dogmatic <- cbind(coefs_dogmatic, robust_se)

@@

stargazer(reg, reg_ego, reg_stereotype, reg_bayesian, reg_dogmatic,  
          title='Regression of effort change on news difference with robust standard errors', 
          dep.var.labels='Change in effort', 
          covariate.labels=c('Good news', 'News difference', 'News difference * Good news'),
          column.labels=c('All', 'Ego-relevant', 'Stereotype', 'Bayesian Simulation', 'Dogmatic Simulation'), 
          omit.stat=c('f', 'ser'), 
          digits=2, 
          label='tab:regression', 
          header=FALSE,
          float.env='sidewaystable', 
          se = list(coefs[, 2], coefs_ego[, 2], coefs_stereotype[, 2], coefs_bayesian[, 2], coefs_dogmatic[, 2]))

### 2023-10-11 Paper/draft.tex
intro start

\setlength{\parindent}{25pt}
\setlength{\parskip}{0pt}
\setlength{\skip\footins}{0.25cm} % margin before footnotes
\definecolor{myblue}{RGB}{31, 61, 122}
\definecolor{mypink}{RGB}{204, 0, 82}
\newtheorem{assumption}{Assumption}
\newtheorem{proposition}{Proposition} % Use the 'proposition' counter for numbering

@@

\title{Learning with misspecified models: Overconfidence and
Stereotypes}
\author{Jimena Galindo}
\date{October 09, 2023}

@@

Overconfidence has been documented in a wide range of settings and has
been shown to have important economic consequences: \citet{Camerer1999}
find that overconfidence can explain the excessive entry of
entrepreneurs into the market; \citet{Hoffman2020} show that workers who
overestimate their productivity within a firm are less likely to look
for a new job; \citet{Oster2013} determine that people at high risk of
Huntington's disease make saving decisions as if they were not at risk.

Although the bias has been widely studied by economists, the settings in
which the the belief updating procedure has been studied is narrow in
scope. In particular, the literature has focused on the cases in which
the agent updates their beliefs about a single characteristic that they
might care about. For instance, \citet{Eil2011} study the case in which
the agent updates their beliefs about their own ability, and their own
beauty in two separate experiments. Similarly, \citet{Benoit2015} study
the case in which the agent updates their beliefs about their own IQ.

@@

I recruited 0m2xzxgv, 0zs7j0mw, 13f4qqvo, 1c7v4nh9, 1gu1vzgk, 1yeajno3,
28arcsnx, 32k6klan, 49lqkk00, 72o8c5bk, 73pn8n63, 78jeza78, 7ebbfqqg,
7h604tn3, 83tbxcch, 8yvmba6x, 9lzz0jai, 9om8jvev, 9rpnegmp, 9w15nflu,
bj048c8i, d37p2ydy, dxfg6c2i, e44vpjp5, fpvdnkq8, g2yvlu4k, hp19zqeh,
inak0r18, iomh77xp, j6ftwuzj, nxd05o0f, p1b4dbgv, p1k5rpou, pxz3760k,
q9icrsb2, qoio0y3d, rp5nv7ut, sfi4nrq8, su9at1s5, szq917jn, vybvjxs2,
xvmd7coq, y71ca1at, z6bycrw5, zap4p51l, 0kxsrg73, 1zb27tea, 36b7nxpg,
3cv0gsso, 4d7lvql9, 4kecnisx, 56op1uib, 6sybrxjn, 6zd6yym2, 78sdphga,
a22g0du7, byf6tjt0, cioxyqvq, cl0818oa, dk93ewke, ffd2bxk4, g2p53iqv,
idzf7yro, ilooy3i8, j3r6tuv8, jk70axat, jka2nr9v, k43107gm, ko3jhnbj,
lmiiifw0, n86dcmzu, ncvtgyzn, oqf1d07l, q9jar1of, uz5m20q1, v3yunf9n,
wursesku, zwzznpo6 undergraduate subjects from the CESS lab at NYU who
participated in an in-person experiment. Sessions lasted approximately
45 minutes and subjects earned an average payment of \(\$22.6\). The
experiment was programmed using oTree \citep{otree}.

@@

So far I have accounted for 21\% subjects as being trapped, and 39\%
having learned the state correctly. From the remaining 35\%, 60\% were
prone to traps but did not fall into them, therefore they could not have
been dogmatic modelers. Alternative explanations are that they were
switchers who considered an incorrect alternative model, or that they
were updating with a bias so large that their effort choices were far
from optimal. Behavior that is not consistent with either of these falls
outside of the scope of the theories discussed above.

@@

\begin{table}[!htbp] \centering 
  \caption{Regression of effort change on news differential with robust standard errors} 
  \label{tab:regression} 
\begin{tabular}{@{\extracolsep{5pt}}lcccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{4}{c}{\textit{Dependent variable:}} \\ 
\cline{2-5} 
\\[-1.8ex] & \multicolumn{4}{c}{Change in effort} \\ 
 & All & Ego-relevant & Stereotype & Bayesian Simulation \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4)\\ 
\hline \\[-1.8ex] 
 Good News & $-$0.12$^{**}$ & $-$0.16$^{***}$ & $-$0.05 & 0.08 \\ 
  & (0.05) & (0.05) & (0.05) & (0.05) \\ 
  & & & & \\ 
 News differential & 0.22$^{***}$ & 0.22$^{***}$ & 0.21$^{***}$ & 0.06$^{***}$ \\ 
  & (0.02) & (0.02) & (0.02) & (0.02) \\ 
  & & & & \\ 
 News Differential * Good News & $-$0.27$^{***}$ & $-$0.25$^{***}$ & $-$0.29$^{***}$ & $-$0.04 \\ 
  & (0.02) & (0.02) & (0.02) & (0.02) \\ 
  & & & & \\ 
 Constant & 0.31$^{***}$ & 0.31$^{***}$ & 0.30$^{***}$ & $-$0.08$^{*}$ \\ 
  & (0.04) & (0.04) & (0.04) & (0.04) \\ 
  & & & & \\ 
\hline \\[-1.8ex] 
Observations & 4,680 & 2,700 & 1,980 & 4,680 \\ 
R$^{2}$ & 0.04 & 0.04 & 0.04 & 0.05 \\ 
Adjusted R$^{2}$ & 0.04 & 0.04 & 0.04 & 0.05 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table}

@@

@@

\renewcommand\refname{References}
  \bibliography{references.bib}

>>>>

\setlength{\parindent}{25pt}
\setlength{\parskip}{0pt}
\setlength{\skip\footins}{0.25cm} % margin before footnotes
\definecolor{myblue}{RGB}{31, 61, 122}
\definecolor{mypink}{RGB}{204, 0, 82}
\newtheorem{assumption}{Assumption}
\newtheorem{proposition}{Proposition} % Use the 'proposition' counter for numbering
\usepackage{rotating, graphicx}

@@

\title{Learning with misspecified models: Overconfidence and
Stereotypes}
\author{Jimena Galindo}
\date{October 10, 2023}

@@

A growing body of literature in economics studies the mechanisms through
which economic agents come to hold incorrect beliefs about a
fundamental. Most of this literature focuses on the case in which agents
are passive observers of the world and incorrectly incorporate the
information they receive into their
beliefs\footnote{See \citet{benjamin2019} for a review of error in probabilistic thinking}.
However, in many situations, agents are active participants in the
process of generating information. In such cases, the information they
receive is affected by their actions and the way they process it can
affect their future behavior. I refer to this as an
\emph{endogenous information} setting. This sort of setting, has been
shown to explain pehnomena that cannot be rationalized with the
classical models of biased belief updating (e.g. \citet{Nyarko1991} and
\citet{Levy2022}). Settings with endogenous information and misspecified
learning also offer alternative explanations for why certain well known
biases might persist. I focus on a group of theories that aim to explain
the prevalence of overconfidence.

Overconfidence has been shown to be a prevalent bias in a variety of
settongs. \citet{Oster2013} shows that subjects who are at high risk of
having Huntingtons disease underestimate their probability of having the
disease and make retirement decisions as if they were not sick. Relating
to investment decisions, \citet{Hoffman2020} show that workers
overestimate their productivity in their current employment and are
unlikely to look for other oportunities. \citet{Camerer1999} find that
entrepreneurs are overconfident about the quiality of their enterprise,
which leads to excessive entry and early exit from markets. These are
only some examples of how holding an incorrect belief about a
fundamental can lead to costly decisions \footnote{Other examples 
of overconfidence are }.

I isolate the main forces from 3 different theories of misspecified
learning. Each of the theories rationalizes the prevalence of
overconfidence in frameworks where the agents are learning about an
ego-relevant parameter and an exogenously determined state. The
interaction between these two parameters together with an endogenous
information process give rise to two explanaitions for the presistence
of overconfidence. On one hand, even with bayesian updating, the agent
may fall into learning traps. Learning traps are misspecified beliefs
and optimal actions that induce signals that confirm the misspecified
belief. Once an agent finds themselves in such a situation, the belief
will be stable and even with a correctly speficified model of learning,
they will not be able to abandon their icorrect beliefs. On the other
hand a setting with two variables allows the agent to engage in
asymmetric updating with respect to good and bad news. When the outcome
depends on an ego-relevant parameter as well as a state, the agent can
attribute good news to a high value of the parameter while attributing
failures to a low value of the exogenous
parameter\footnote{This is in line with the well known theory 
in behavioral science of self-attribution bias of which \citet{kelley1980} provides a review.}.

\citet{Heidhues2018} shows that under certain assumptions on the output
function, learning traps can rationalize the existance of misspecified
beliefs when agents are dogmatic about the ego-relevant parameter. In a
more general model \citet{Ba2023} further analyzes this setting and
characterizes the stuations in which even agents who are not dogmatic
might become trapped. She allows initi dogmatic agents to consider
alternative theories and adopt whichever is a better explanation of the
observed data. They do so by performing Bayesian hypothesis tests that
compare the likelihood of having observed the outcome under alternative
paradigms\footnote{A 
similar mechanism is proposed by \citet{Schwarstein2021} in a setting with persuasion}.
Although this learning mechanism allows agents to avoid the learning
traps, there are still situations under which misspecified beliefs may
persist.

@@

I recruited 78 undergraduate subjects from the CESS lab at NYU who
participated in an in-person experiment. Sessions lasted approximately
45 minutes and subjects earned an average payment of \(\$22.6\). The
experiment was programmed using oTree \citep{otree}.

@@

So far I have accounted for 16\% subjects as being trapped, and 39\%
having learned the state correctly. From the remaining 45\%, 60\% were
prone to traps but did not fall into them, therefore they could not have
been dogmatic modelers. Alternative explanations are that they were
switchers who considered an incorrect alternative model, or that they
were updating with a bias so large that their effort choices were far
from optimal. Behavior that is not consistent with either of these falls
outside of the scope of the theories discussed above.

@@

\begin{sidewaystable}[!htbp] \centering 
  \caption{Regression of effort change on news difference with robust standard errors} 
  \label{tab:regression} 
\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{5}{c}{\textit{Dependent variable:}} \\ 
\cline{2-6} 
\\[-1.8ex] & \multicolumn{5}{c}{Change in effort} \\ 
 & All & Ego-relevant & Stereotype & Bayesian Simulation & Dogmatic Simulation \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4) & (5)\\ 
\hline \\[-1.8ex] 
 Good news & $-$0.12$^{**}$ & $-$0.16$^{***}$ & $-$0.05 & 0.08 & $-$0.08 \\ 
  & (0.05) & (0.05) & (0.05) & (0.05) & (0.05) \\ 
  & & & & & \\ 
 News difference & 0.22$^{***}$ & 0.22$^{***}$ & 0.21$^{***}$ & 0.06$^{***}$ & 0.10$^{***}$ \\ 
  & (0.02) & (0.02) & (0.02) & (0.02) & (0.02) \\ 
  & & & & & \\ 
 News difference * Good news & $-$0.27$^{***}$ & $-$0.25$^{***}$ & $-$0.29$^{***}$ & $-$0.04 & $-$0.06$^{***}$ \\ 
  & (0.02) & (0.02) & (0.02) & (0.02) & (0.02) \\ 
  & & & & & \\ 
 Constant & 0.31$^{***}$ & 0.31$^{***}$ & 0.30$^{***}$ & $-$0.08$^{*}$ & 0.05 \\ 
  & (0.04) & (0.04) & (0.04) & (0.04) & (0.04) \\ 
  & & & & & \\ 
\hline \\[-1.8ex] 
Observations & 4,680 & 2,700 & 1,980 & 4,680 & 4,680 \\ 
R$^{2}$ & 0.04 & 0.04 & 0.04 & 0.05 & 0.06 \\ 
Adjusted R$^{2}$ & 0.04 & 0.04 & 0.04 & 0.05 & 0.06 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{5}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{sidewaystable}

@@

\hypertarget{references}{%
\section{References}\label{references}}

@@

  \bibliography{references.bib}
